A generalized definition of multidimensional item response theory parameters

### 2024-07-15 output/paper_draft/paper_M2PLM_parameters.Rmd
Improves discussion after latest results

Multidimensional item response theory (MIRT) models have been widely used since
their inception.
The so-called *Multidimensional 2-Parameter Logistic*
[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is among the most important
and widely used ones.
This is partly due to its close relationship to the common factor model
[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
It consists of a generalization of the 2-parameter logistic model
[@birnbaum_latent_1968] to more than one latent trait
[@mckinley_extension_1983].
Its usefulness lies in describing the variability of the population when
the correct response does not depend on a single trait;
more specifically, when a deficit in a certain trait
and a surplus in another one cancel out their effects. Because of this property,
it is sometimes referred to as a *compensatory* model.

@reckase_difficulty_1985 argues for a need to take dimensionality into account
when characterizing the statistical properties of multidimensional items.
As such, he outlines a general procedure for determining their difficulty.
Following this procedure, he also derives
a *multidimensional difficulty* parameter (MID) for `r MODEL_ACRONYM` items.
This parameter consists of two parts:
The first one is a scalar value representing the distance from
the origin of the multidimensional latent space to
the maximally discriminating locus of the item;
the second one, the direction of the vector from the origin to that same locus.
Following a similar rationale, @reckase_discriminating_1991 derive a
*multidimensional discrimination* (`r MDISC_SYM`) parameter definition
for that model.

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
@reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
---exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
the R package *psych* [@revelle_psych_2024], graphically represents
factor loadings on orthogonal axes even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
This fact is probably due to the different spotlights
these two theoretical programs put the focus on;
factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] requires
the relaxation of traditional constraints on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration
[see @zhang_theoretical_1999 for a notable exception].
Moreover, disregarding this geometrical interpretation may lead to
critical misconceptions about the measurement model.

The representation of a latent space where the traits are correlated is more
accurate on a geometrical coordinate system with non-orthogonal axes and
non-standard units, often referred to as
*general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the basis vectors and the axes they yield
have a meaningful interpretation (albeit counterintuitive, as we will see)
in a multivariate statistical sense.
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
[e.g., @thielmann_comparability_2022].
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait, and thus
usually reverse-scored--- is also widespread
[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
of the Big-Five Factor Markers, @noauthor_international_nodate].
It is common to find instruments that combine
positively- and negatively-keyed items.
Moreover, a complex item may have discrimination parameters with opposite signs
for different latent dimensions.<!--# Examples? -->
Take the multidimensional version of the nominal categories model
[@thissen_nominal_2010] as an example,
and consider two response categories tapping a different construct each:
A positive change in the dimension tapped by one category will inevitably
lead to a decrease in the probability of response to the other category.
One can easily note that a monotonically decreasing probability for a certain
dimension is given by a non-positive discrimination parameter
(and an unchanging probability by a null discrimination parameter,
which also applies to cognitive instruments),
contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
However, the multidimensional item parameters can also be defined for
items with monotonically non-increasing probabilities, as we will see.

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim to provide a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application,
which will require introducing some linear algebra concepts and results.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce three different versions,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss their properties,
propose a graphical representation method based on the new formulation,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based index has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
Once again, this is important in applications where
potentially different latent spaces are involved.
Given these considerations, we propose to adopt the covariance-based version of
the multidimensional parameters and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of preserving the metric of the parameters,
which is convenient for fidelity when representing the items.
Therefore, we recommend using the correlation-based version of the parameters
for graphical representation purposes only.

One may consider that the formal derivations presented here are overly
complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Moreover, strictly following @reckase_difficulty_1985's method in the original,
oblique space has also provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, in order to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although models such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying measurement theory, its implications and its interpretation.

>>>>

Multidimensional item response theory (MIRT) models have been widely used since
their inception.<!--# Ref. needed? -->
The so-called *Multidimensional 2-Parameter Logistic*
[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is among the most important
and widely used ones.
This is partly due to its close relationship to the common factor model
[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
It consists of a generalization of the 2-parameter logistic model
[@birnbaum_latent_1968] to more than one latent trait
[@mckinley_extension_1983].
Its usefulness lies in describing the variability of the population when
the correct response does not depend on a single trait;
more specifically, when a deficit in a certain trait
and a surplus in another one cancel out their effects. Because of this property,
it is sometimes referred to as a *compensatory* model.

@reckase_difficulty_1985 argues for a need to take dimensionality into account
when characterizing the statistical properties of multidimensional items.
As such, he outlines a general procedure for determining their difficulty.
Following this procedure, he also derives
a *multidimensional difficulty* parameter (MID) for `r MODEL_ACRONYM` items.
This parameter consists of two parts:
The first one is a scalar value representing the distance from
the origin of the multidimensional latent space to
the maximally discriminating locus of the item.
The second one, the direction of the vector from the origin to that same locus.
Following a similar rationale, @reckase_discriminating_1991 derive a
*multidimensional discrimination* (`r MDISC_SYM`) parameter definition
for that model.

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
@reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
---exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
the R package *psych*, <!--# Reference needed? -->graphically represents
factor loadings on orthogonal axes even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
This fact is probably due to the different spotlights
these two theoretical programs put the focus on;
factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] requires
the relaxation of traditional constraints on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
[see @zhang_theoretical_1999 for a notable exception].
Moreover, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

A latent space where the traits are correlated is easier to understand as a
geometrical space with non-orthogonal axes,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the cosines of the angles between two axes
can be interpreted as the correlation of the corresponding latent dimensions.
<!--# Reference needed? -->
<!-- Problem here: The latent space is not defined like that, according to
the results of section "Linear algebra of non-orthogonal coordinates". -->
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait, and thus
usually reverse scored--- is also widespread.<!--# Reference and example? -->
It is common to find instruments that combine
positively- and negatively-keyed items.
Moreover, a complex item may have discrimination parameters with opposite signs
for different latent dimensions.<!--# Examples? -->
Take the multidimensional version of the nominal categories model
[@thissen_nominal_2010] as an example,
and consider two response categories tapping a different construct each:
A positive change in the dimension tapped by one category will inevitably
lead to a decrease in the probability of response to the other category.
One can easily note that a monotonically decreasing probability for a certain
dimension is given by a non-positive discrimination parameter
(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
However, the multidimensional item parameters can also be defined for
items with monotonically non-increasing probabilities, as we will see.

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim at providing a method for
graphically representing these items in the fashion of @ackerman_creating_1994
but taking the non-orthogonality of the coordinate axes into account.
In order to pursue these objectives, we first define the generalization of
the `r MODEL_ACRONYM` model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next, we discuss the properties of the indices,
propose a graphical representation method,
and present two examples of application of our results.
We round up with a discussion about
the applicability and generalization of our results.

@@

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based index has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is important in applications where
potentially different bases are involved.
Given these considerations,
we propose to adopt the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, it is worth noting that it has been defined ad-hoc
to fulfill this specific property.
Therefore it lacks other desirable properties of the covariance-based version,
like being invariant to a latent space scaling transformation,
or yielding a diagonal latent covariance matrix in the corresponding
orthogonal latent space basis.
Therefore, we recommend its use for graphical representation purposes only.
As for the agnostic version, we think it should only be used in
the (unlikely) case of having no information about the latent space structure.

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of the parameters to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require to compute the cosines in
the original basis afterwards.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, in order to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
This result is both surprising and counterintuitive;
The MIRT literature suggests in several instances that
"the angle between the axes [of latent ability space] represents
the degree of correlation" [@, p. 16].
We hope our results help shed a clearer light on
this often missunderstood topic.

It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although proposals such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying multidimensional measurement theory, its implications,
and its interpretation.
